---
title: "ostrichtermination_Test"
date: "2026-08-26"
output: 
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#**Question: Does soil moisture and distance of tree from water have an effect on moss cover on Trembling Aspen at North and South ASCCA? 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
# Loading the dataset
mossdata <- read_delim("moss_data.csv", delim = ',')
#Renaming the dataset to mossdata

```
#**Mean Moss cover vs soil moisture across all four sites combined: 

#Should we use ggplot here for consistency? I think it works either way
```{r}
plot(mossdata$`Soil Moisture (unit)`, 
     mossdata$`% cover`,
     xlab = "Soil Moisture",
     ylab = "Moss cover (%)" )
```
#The data seems to be skewed to the right. 
```{r}
shapiro.test(mossdata$`% cover`)
```
#The data is NOT normally distributed according to the shapiro-wilk normality test. 
```{r}
library(fitdistrplus)
descdist(mossdata$`% cover`)
```

#Cullen and Frey graph suggest gamma distribution 

#need to use gamma family for glm, do not forget to omit N/As when using distance from water in models. Look for AIC values of %cover * soil moisture, %cover * zone, %cover * distance from water, and a final model with all interactions. Also look at relationship between distance from water and soil moisture (because some areas were observed to be dryer or wetter than expected). 

#Cleaning the data and renaming/creating labels, while "N/A" values are treated as missing.
```{r}

mossdata <- read_csv("moss_data.csv", na = c("N/A", "NA", ""))

moss_clean <- mossdata %>%
  rename(
    date = Date,
    zone = Zone,
    tree_id = `Tree ID`,
    distance_water = `Distance from water (m)`,
    soil_moisture = `Soil Moisture (unit)`,
    moss_height = `Moss Height (cm)`,
    moss_width = `Moss Width (cm)`,
    bhd = `bhd (cm)`,
    moss_cover = `% cover`,
    temperature = `Temperature (°C)`
  )

summary(moss_clean)

```

#Making a water distance subset
```{r}

moss_distance <- moss_clean %>%
  drop_na(distance_water)

# Check the number of observations in each dataset.
nrow(moss_clean)
nrow(moss_distance)

```

#Exploratory varaibles and graphs + soil moisture and moss cover across all sites
```{r}

ggplot(moss_clean, aes(x = soil_moisture, y = moss_cover)) +
  geom_point() +
  labs(
    x = "Soil Moisture",
    y = "Moss Cover (%)"
  ) +
  theme_classic()

```
#Analysis of above: 


#Distribution of moss cover
```{r}

ggplot(moss_clean, aes(x = moss_cover)) +
  geom_histogram() +
  labs(
    x = "Moss Cover (%)",
    y = "Frequency"
  ) +
  theme_classic()

```
#Analysis of above:

#Soil moisture and moss cover
```{r} 

model_soil <- glm(
  moss_cover ~ soil_moisture,
  data = moss_clean,
  family = Gamma(link = "log")
)

summary(model_soil)
AIC(model_soil)

```
#Visualizing this model throuhg a scatterplot
```{r}
ggplot(moss_clean, aes(x = soil_moisture, y = moss_cover)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list(family = Gamma(link = "log"))) +
  labs(
    x = "Soil Moisture",
    y = "Moss Cover (%)"
  ) +
  theme_classic()

```

#Now we need to see what are the differences in moss coverage among each zone
```{r}

model_zone <- glm(
  moss_cover ~ zone,
  data = moss_clean,
  family = Gamma(link = "log")
)

summary(model_zone)
AIC(model_zone)

```

#Visualizing it again

```{r}

ggplot(moss_clean, aes(x = zone, y = moss_cover)) +
  geom_boxplot() +
  labs(
    x = "Sampling Zone",
    y = "Moss Cover (%)"
  ) +
  theme_classic()

```

# Soil moisture + zone model
```{r}

model_soil_zone <- glm(
  moss_cover ~ soil_moisture + zone,
  data = moss_clean,
  family = Gamma(link = "log")
)

summary(model_soil_zone)
AIC(model_soil_zone)

```

#Soil moisture and zone interaction model
```{r}

model_soil_zone_int <- glm(
  moss_cover ~ soil_moisture * zone,
  data = moss_clean,
  family = Gamma(link = "log")
)

summary(model_soil_zone_int)
AIC(model_soil_zone_int)

```

#Now we need to compare all of the models using AIC's

```{r} 

AIC(
  model_soil,
  model_zone,
  model_soil_zone,
  model_soil_zone_int)
  
```
#Interpertaion: 

#Now we need the distance from water and moss cover
```{r}

model_distance <- glm(
  moss_cover ~ distance_water,
  data = moss_distance,
  family = Gamma(link = "log")
)

summary(model_distance)
AIC(model_distance)
```

#Another gamma model to visualize

```{r}

ggplot(moss_distance,
       aes(x = distance_water, y = moss_cover)) +
  geom_point() +
  geom_smooth(
    method = "glm",
    method.args = list(family = Gamma(link = "log"))
  ) +
  labs(
    x = "Distance From Water (m)",
    y = "Moss Cover (%)"
  ) +
  theme_classic()

```
#Interpertaion

#Combined distance model containing soil moisture, distance from water, and zone to investigate moss cover
```{r}

model_distance_full <- glm(
  moss_cover ~ soil_moisture + distance_water + zone,
  data = moss_distance,
  family = Gamma(link = "log"))

summary(model_distance_full)
AIC(model_distance_full)

```

#almost there but now we wanted to test soil moisture and distance from water
```{r} 

model_distance_int <- glm(
  moss_cover ~ soil_moisture * distance_water + zone,
  data = moss_distance,
  family = Gamma(link = "log")
)

summary(model_distance_int)
AIC(model_distance_int)

```
#results


#testing AICs again
```{r}

AIC(
  model_distance,
  model_distance_full,
  model_distance_int)

```
#results

#relationship between soil mouisture and distance from water
```{r}

model_moisture_distance <- lm(
  soil_moisture ~ distance_water,
  data = moss_distance
)

summary(model_moisture_distance)
plot(model_moisture_distance)

```

#visualize again

```{r}
ggplot(moss_distance, aes(x = distance_water, y = soil_moisture)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Distance From Water (m)",
    y = "Soil Moisture"
  ) +
  theme_classic()

```

#Summary of statistical approach 

