---
title: "ostrichtermination_Test"
output: html_document
date: "2026-08-26"
output: 
  pdf_document:
    toc: yes
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#**Question: Does soil moisture and distance of tree from water have an effect on moss cover on Trembling Aspen at North and South ASCCA? 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
# Loading the dataset
mossdata <- read_delim("moss_data.csv", delim = ',')
#Renaming the dataset to mossdata

```
#**Mean Moss cover vs soil moisture across all four sites combined: 
```{r}
plot(mossdata$`Soil Moisture (unit)`, 
     mossdata$`% cover`,
     xlab = "Soil Moisture",
     ylab = "Moss cover (%)" )
```
#The data seems to be skewed to the right. 
```{r}
shapiro.test(mossdata$`% cover`)
```
#The data is NOT normally distributed according to the shapiro-wilk normality test. 
```{r}
library(fitdistrplus)
descdist(mossdata$`% cover`)
```

#Cullen and Frey graph suggest gamma distribution 

#need to use gamma family for glm, do not forget to omit N/As when using distance from water in models. Look for AIC values of %cover * soil moisture, %cover * zone, %cover * distance from water, and a final model with all interactions. Also look at relationship between distance from water and soil moisture (because some areas were observed to be dryer or wetter than expected). 

